Software architectures to integrate workflow engines in science gateways

### 2016-05-15 paper.tex
Removed closed comments as they may introduce white spaces.

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\closedanswerednote}[6]{}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed
\closednote{vladimir}{is this a good word for this context,
  'informed'? maybe 'based on'?}{Tristan}{I guess it is the right
  word, keeping it as is.} by our experience in the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years, as well as by lessons learned from several science
gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
provide background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures
\closednote{vladimir}{why is it exactly six? are these architectures
  the ones that happened to be implemented during various projects, or
  is it a full set of all possibilities (I believe not)? I think, it
  must be stated clearly why the focus is given to these particular
  ones - we are reviewing and evaluating the approaches that have been
  used already in our projects}{Tristan}{Fixed, thanks.} that cover
the main patterns that we encountered in the various systems that we
studied during the last few years. The architectures are described
within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. \closednote{vladimir}{can we explain why exactly these
  metrics are used but not anything else? I think, we need to justify
  that these ones are especially important for the SG case; it would
  be nice to refer to some related research on software architecture
  metrics, Google gives about 2 million pages on this, where you can
  find metrics like cohesion, depth of scenario and so
  on}{Tristan}{The metrics are specifically designed to cover our
  particular workflow-SG context. Do you have any other
  suggestion?} We finally compare the architectures and highlight
recommendations for science gateway architects and 
developers.

@@

\label{sec:introduction-engines}
In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}. \closednote{vladimir}{As a past developer of it, I
  would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it
  is not very famous, but it played an important role for Amsterdam
  e-science workflows}{Tristan}{Done, thanks.} Such workflow engines usually describe
applications in a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

  may require specific privileges; \closednote{vladimir}{or bundled in
    packages considering all needed dependencies than can be deployed
    on the fly during the workflow execution}{Tristan}{added, thanks!} (\texttt{a$_2$}) the

@@

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, as a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g., RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although in some
variations of the architecture (not represented in
Figure~\ref{fig:architectures}) it might as well be delivered by the
workflow engine directly to the user. \closednote{Sil}{about c1: should it
  also show from engine to infra in the figure? this is what i
  understand from the above sentence}{Tristan}{Actually c1 is either
  coming from the science gatewray or from the workflow engine but
  then it merges with c2. I don't think we can cover both cases in a
  single architecture, the latter case would be another variation of
  this architecture. I modified a bit the text to reflect this.}

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflow engines are integrated in the science gateway
through workflow format conversion from a native format to the science
gateway format. \closednote{vladimir}{Are we talking only about importing
  workflows from other wf languages? I would suggest to modify the
  title of the section to 'Workflow conversion and import' then, as
  some gateways use the term 'import' to import their own native
  workflows, like WS-PGRADE, and using import only in the context of
  converting workflows can be confusing}{Tristan}{Good catch, the term
  'import' is actually a bit vague. I guess 'conversion' is more
  accurate. Replaced it everywhere in the text.} Workflow conversion
is usually an offline process that is not involved in workflow
execution. A few systems have implemented workflow conversion. In the
SHIWA Simulation Platform, it is implemented through the IWIR
language, which provides a common language for portability across grid
workflow systems~\cite{plankensteiner-prodan-etal:2013} and allows
conversion among $n$ workflow languages using $2n$ interactions
instead of $n^2$.  Tavaxy~\cite{Abouelhoda2012} enables the import,
merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow conversion from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE
and from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
\closednote{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered. One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly. So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}{Tristan}{In our current evaluation framework, this is supposed to be covered by O2. I still agree with you that O2 could be further refined. See Github issue #27}
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  \closednote{vladimir}{is this the right word here, 'to
    means'?}{Tristan}{This sentence had been truncated, thanks.} that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  amd \texttt{O$_1$}=1 means that they may be implemented with some
  development effort. \closednote{si}{why here we do not have the
    "possible to implement" situation?}{Tristan}{O$_1$=2 would never
    be reached. Besides, that would make it difficult to sum O1 and O2
    because they wouldn't measure the same type of thing. In the
    current situation, we are only talking about the presence/absence
    of a particular feature and this is why O1 and O2 can be added to
    obtain a total number of 'missing features'.}
  \closednote{vladimir}{I agree with Silvia, we should have the
    intermediate values as well, since meta-workflows might be
    straightforwardly supported by the gateway, or they are still
    possible to implement but only with some workarounds. Like it was
    with SHIWA: first 6 months coarsed-grained interoperability looked
    more like a workaround till it was better integrated and supported
    by WS-PGRADE itself}{Tristan}{Yes but in the situation you
    describe, evolutions came from the wf engine implementation rather
    than from the architecture to integrate it, which is our point
    here.}

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow).  Fine-grained information about workflow tasks is
  required to properly troubleshoot workflow executions. Its
  availability and accuracy usually depends on the number of software
  layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that
  the information is directly accessible in the science gateway,
  \texttt{O$_2$}=1 means that it is obtained through one or more software
  interactions. \closednote{vladimir}{otherwise - does it mean the
    fine-grained info is available but hard to access or not available
    it all? the latter happens as well}{Tristan}{Precised the text.}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

\closednote{sil}{since we are dreaming, it seems to me that keeping provenance properly is also something to consider. the architecture can hide some details, and make it nearly impossible to keep provenance without hard coded hacks across the layers/components - i think}{Tristan}{Fixed the definition according to the discussion in PR #20.}

@@

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the complete science gateway
%, which is in general not possible
\closednote{sil}{why not? i agree that S=2, but i do not agree that it would be impossible. you say how below}{Tristan}{Updated the definition of S$_1$ to explain that multiple engine instances have to be available in the same sg instance.}
(\texttt{S$_1$}=2). 
Specific IT setups such as load-balancing between
web server instances might be used to create new gateway instances, but they would not allow a single gateway instance to use multiple engine instances, which is the point here.
Distributed engines are not available by default
(\texttt{S$_2$}=1). 
The architecture does not add any particular complexity to the task scheduling problem since the workflow engine
may implement any kind of scheduling policy (\texttt{S$_3$}=0).
\closednote{sil}{so same as the others = "easy"? i think this requires more explanation. maybe we should introduce a score for "dont know"?}{Tristan}{Fixed as discussed in PR #20}

\paragraph{Specific features} Supporting meta-workflows requires
specific developments in the workflow engine or science gateway to
allow workflows to execute other workflows (\texttt{O$_1$}=1). For
instance, specific Java objects may be implemented in the science
gateway framework to execute different types of workflows and link
them together.  \closednote{sil}{why not? if the engine is ws-pgrade
  it does. i really think that meta-workflows (and nested workflows)
  is a property exclusively of the engine. you can see that in fig 2.V
  interaction *6}{Tristan}{If the engine is WS-PGRADE as used in SHIWA
  CGI, then the architecture is nested workflows with service
  invocation. In a tight integration, meta-workflows would be
  expressed, e.g. as compositions of Liferay portlets that all
  implement different types of engines.}  The science gateway can
retrieve fine-grained debugging information from the workflow engine
directly (\texttt{O$_2$}=0).

@@

\paragraph{Scalability} The service architecture in principle supports
multiple engine instances through multiple workflow services. Adding a
new engine instance, however, requires specific developments in the science gateway
(\texttt{S$_1$}=1). Distributing the execution of a single workflow in multiple
engines is usually not possible unless the workflow engine has
specific abilities (\texttt{S$_2$}=1).  \closednote{sil}{is this score for
  VIP? i think that the score depends on the implementation of the
  science gateway, which could adopt some elastic mechanism within. so
  we should not say s_2=1 for the architecture}{Tristan}{Removed VIP example. Updated score definition according to the discussion in @SilviaOlabarriaga's PR (\#20).} The architecture does not add any particular complexity to the task scheduling problem
since the workflow engine may implement any kind of scheduling
policy (\texttt{S$_3$}=0).  \closednote{sil}{see comment in section 3.1}{Tristan}{Fixed the same way}

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine or science gateway to allow
workflows to execute other workflows.  (\texttt{O$_1$}=1).
\closednote{sil}{see comment in section 3.1}{Tristan}{same answer} The
science gateway needs to invoke interaction \texttt{d} to retrieve
dine-grained debugging information from the workflow service
(\texttt{O$_2$}=1).  \closednote{sil}{this is really not the case. the
  fact that the wf service is a separate component that hides
  information is a big problem for debugging at the science gateway
  level.  we had a paper about this:
  @inproceedings{olabarriaga2014scientific, Author = {Olabarriaga,
      Silvia and Pierantoni, Gabrielle and Taffoni, Giuliano and
      Sciacca, Eva and Jaghoori, Mahdi and Korkhov, Vladimir and
      Castelli, Giuliano and Vuerli, Claudio and Becciani, Ugo and
      Carley, Eoin and others}, Booktitle = {e-Science (e-Science),
      2014 IEEE 10th International Conference on}, Organization =
    {IEEE}, Pages = {298--305}, Title = {Scientific Workflow
      Management--For Whom?}, Year = {2014}} }{Tristan}{I agree. Fixed
  it and precised the definition of O2.}

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is necessary
(\texttt{R$_2$}=1).
\closednote{vladimir}{there is a large piece of text in Scalability about reliability of the workflow engine task which on my opinion suits Robustness better}{Tristan}{I understand that that text may sound like robustness, but the robustness metrics are about the architecture, not the tasks, as explained in 3.1 (Errors coming from
the infrastructure (e.g., unavailable data or termi-
nated tasks) or workflows (e.g., wrong user input or
application errors) are not covered).}
\closednote{vladimir}{in addition, the claim of very good robustness of task encapsulation seems questionable to me as in this model you delegate the critical task of controlling workflow execution to, let's say, some 'random' node on the grid that you have no control of, not sure about its performance, connectivity and reliability and so on; however, considering the models and dependencies presented in fig.2 this evaluation of robustness metric may be correct}{Tristan}{You are right but the robustness issues that you are talking about are related to the infrastructure and not the architecture. This is really an issue related to task scheduling as the SG might decide to run the workflow tasks on reliable nodes of the infrastructure such as a dedicated cluster. So the issue that you mention is reflected in S3.}

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). 
Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. \closednote{vladimir}{isn't this about robustness?}{Tristan}{Nope, this is task scheduling, therefore scalability in our evaluation framework. See answer to your previous note.} In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen.

\paragraph{Specific features} Task encapsulation explicitly enables
meta-workflows (\texttt{O$_1$}=0). Indeed, workflow engines may invoke
other engines using the exact same interaction used to submit regular
sub-tasks (interaction \texttt{e}), without any additional
development. \closednote{sil}{see comment in section
  3.1}{Tristan}{Fixed, actually updated the score of this architecture
  now that the definition is more accurate.} Fine-grained debugging
information is obtained through interaction \texttt{b}
(\texttt{O$_2$}=1). \closednote{sil}{mm, i think the case here is the
  same as for service. or actually i would give this one a better
  score, because there are less components to cross}{Tristan}{Fixed the text to better match the updated definition of O2}

@@

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine to enable workflow submission
(\texttt{O$_1$=1}).  \closednote{sil}{see comment in section 3.1}{Tristan}{Same answer.} Debugging information is accessed through interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).
\closednote{sil}{i think the case here is the same as for service. or
  actually i would give this one a worse score, because there is one
  additional component - the pool - to cross}{Tristan}{Fixed using the updated definition of O2}

@@

\paragraph{Integration complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\closednote{vladimir}{and if we have a meta-workflow with 3 and more sub-workflows the score will be even higher, is it a fair calculation for an architecture? we did not increase the scores for possible meta-workflows supported by wf engines in other cases}{Tristan}{Good point! You are totally right for meta-workflows, I added a sentence in the paragraph below. However, nested workflows are studied here as an architecture to integrate a workflow engine (and only one), not to describe meta-workflows.}

@@

\closednote{vladimir}{@silvia, does it make sense to mention workflow classification efforts based on workflow motifs and so on?}{Tristan}{I'm  not  sure. Which papers are you thinking of?}

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed
 by our experience in the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years, as well as by lessons learned from several science
gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
provide background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures
 that cover
the main patterns that we encountered in the various systems that we
studied during the last few years. The architectures are described
within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. We finally compare the architectures and highlight
recommendations for science gateway architects and 
developers.

@@

\label{sec:introduction-engines}
In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}.  Such workflow engines usually describe
applications in a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

  may require specific privileges; (\texttt{a$_2$}) the

@@

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, as a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g., RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although in some
variations of the architecture (not represented in
Figure~\ref{fig:architectures}) it might as well be delivered by the
workflow engine directly to the user.

@@

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflow engines are integrated in the science gateway
through workflow format conversion from a native format to the science
gateway format. Workflow conversion
is usually an offline process that is not involved in workflow
execution. A few systems have implemented workflow conversion. In the
SHIWA Simulation Platform, it is implemented through the IWIR
language, which provides a common language for portability across grid
workflow systems~\cite{plankensteiner-prodan-etal:2013} and allows
conversion among $n$ workflow languages using $2n$ interactions
instead of $n^2$.  Tavaxy~\cite{Abouelhoda2012} enables the import,
merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow conversion from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE
and from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  amd \texttt{O$_1$}=1 means that they may be implemented with some
  development effort.

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow).  Fine-grained information about workflow tasks is
  required to properly troubleshoot workflow executions. Its
  availability and accuracy usually depends on the number of software
  layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that
  the information is directly accessible in the science gateway,
  \texttt{O$_2$}=1 means that it is obtained through one or more software
  interactions. 
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

@@

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the complete science gateway
%, which is in general not possible
(\texttt{S$_1$}=2). 
Specific IT setups such as load-balancing between
web server instances might be used to create new gateway instances, but they would not allow a single gateway instance to use multiple engine instances, which is the point here.
Distributed engines are not available by default
(\texttt{S$_2$}=1). 
The architecture does not add any particular complexity to the task scheduling problem since the workflow engine
may implement any kind of scheduling policy (\texttt{S$_3$}=0).

\paragraph{Specific features} Supporting meta-workflows requires
specific developments in the workflow engine or science gateway to
allow workflows to execute other workflows (\texttt{O$_1$}=1). For
instance, specific Java objects may be implemented in the science
gateway framework to execute different types of workflows and link
them together.  The science gateway can
retrieve fine-grained debugging information from the workflow engine
directly (\texttt{O$_2$}=0).

@@

\paragraph{Scalability} The service architecture in principle supports
multiple engine instances through multiple workflow services. Adding a
new engine instance, however, requires specific developments in the science gateway
(\texttt{S$_1$}=1). Distributing the execution of a single workflow in multiple
engines is usually not possible unless the workflow engine has
specific abilities (\texttt{S$_2$}=1).

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine or science gateway to allow
workflows to execute other workflows.  (\texttt{O$_1$}=1).
 The
science gateway needs to invoke interaction \texttt{d} to retrieve
dine-grained debugging information from the workflow service
(\texttt{O$_2$}=1).

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is necessary
(\texttt{R$_2$}=1).

@@

\paragraph{Scalability}
New engine instances are spawned and executed on the infrastructure as
any other task upon user submission (\texttt{S$_1$}=0). This is a
major interest of task encapsulation. Distributed engines are not
supported by default (\texttt{S$_2$}=1). Task scheduling is slightly
more complex than in the other approaches due to the special role of
the task that executes the workflow engine (\texttt{S$_3$}=1). 
Indeed,
the reliability of this task is critical since all the sub-tasks in
the workflow depend on it and, depending on the recovery capabilities
of the workflow engine, may need to be resubmitted if the workflow
task fails. The workflow task is also longer than all its sub-tasks,
which increases its chances of failure. In addition, task parameters,
for instance estimated walltime, are more difficult to estimate for
the workflow task than for the sub-tasks because workflows are by
definition more complex than their sub-tasks: errors on sub-task
parameter estimations accumulate in the workflow, and additional
control constructs such as tests and loops may further increase the
uncertainty. Such parameter estimation errors may generate issues such
as selection of wrong batch queues on clusters or task termination due
to exceeded quotas. Finally, the interdependencies between
the workflow task and its sub-tasks may create deadlocks when there is
contention. For instance, if only 1 computing resource is available
for the science gateway and if the workflow task is running on it and
submits sub-tasks, then the sub-tasks could only execute when the
resource is available, which will never happen because the workflow
task will not complete until the sub-tasks complete. This
configuration can be generalized to an infrastructure with $n$
resources where $n$ workflows are submitted. In practice, however, the
number of submitted workflows usually remains lower than the number of
computing resources available on this infrastructure, which makes such
deadlocks unlikely to happen.

\paragraph{Specific features} Task encapsulation explicitly enables
meta-workflows (\texttt{O$_1$}=0). Indeed, workflow engines may invoke
other engines using the exact same interaction used to submit regular
sub-tasks (interaction \texttt{e}), without any additional
development. Fine-grained debugging
information is obtained through interaction \texttt{b}
(\texttt{O$_2$}=1).

@@

\paragraph{Specific features} Meta-workflows require specific
developments in the workflow engine to enable workflow submission
(\texttt{O$_1$=1}). Debugging information is accessed through
interactions \texttt{d} and \texttt{f} (\texttt{O$_2$=1}).

@@

\paragraph{Integration complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).

@@
